Mainstreaming Race Science

### 2023-10-04 paper/paper.Rmd
edits based on comments from K, J, M

acknowledgements: |
  | Thanks to Emily Merchant and John Jackson for some initial discussions that helped clarify the scope of the project and identify some key resources related to the Pioneer Fund and *Mankind Quarterly*.  Thanks to Anthony Sainez for help retrieving the *Mankind Quarterly* articles.  Thanks to Derek Devnich and James Dooley for their work attempting to secure the APA-published articles. 
  | **Funding:** EL's work on this project was partially supported by UC Merced.  DH's work on this project was not supported by any specific funding. 
  | **Author contributions:** DJH (conceptualization, data curation, formal analysis, investigation, methodology, project administration, software, validation, visualization, writing-original draft, writing-review & editing); EL (conceptualization, methodology, software, validation, visualization, writing-original draft, writing-review & editing)
  | **Competing interests:** The authors have no competing interests to declare. 
  | **Data and materials availability:** Web of Science search results and article fulltext are not publicly available due to intellectual property restrictions.  Preprocessed analysis data (document-term counts and article metadata) and analysis scripts are available at *[doi]*. 
# bibliography: "references.bib"
output: 
  # https://github.com/rstudio/rticles/pull/486
  # skeleton: <https://github.com/christopherkenny/rticles/tree/master/inst/rmarkdown/templates/science>
  bookdown::pdf_book:
    base_format: rticles::science_article
    keep_tex: true
    move_figures: false
    draft: true
    number_sections: true
header-includes:
  - \usepackage{lineno}
  - \linenumbers
---

@@

In the landmark case *Brown v Board of Education of Topeka* (1954), the US Supreme Court banned *de jure* educational segregation.  The Court's decision relied on expert testimony from psychologists and education researchers; but the segregationists also put forward their own experts, including Henry Garrett \cite{WinstonScienceServiceFar1998, JacksonScienceSegregationRace2005, SchafferScientificRacismAgain2007}.

As part of the segregationist reaction to *Brown*, MQ was created as what contemporary historians, philosophers, and sociologists of science call an "echo chamber" \cite{FernandezPintoKnowBetterNot2017}, a favorable venue for race scientists to publish their views, on the grounds that an "equalitarian dogma" created a censorious "taboo" against their research in mainstream publications \cite{TuckerFundingScientificRacism2002, JacksonMythicalTabooRace2020}.  Echo chambers are a key part of the "tobacco strategy," used by numerous regulated industries — most infamously tobacco, but also fossil fuels and chemical manufacturing, among others — to "manufacture doubt" and delay regulation \cite{OreskesMerchantsDoubtHow2011, JasnyEmpiricalExaminationEcho2015, FernandezPintoKnowBetterNot2017, WeatherallHowBeatScience2018}.

In light of the origins of Pioneer and MQ, the significant attention paid to the journal by historians of scientific racism, and contemporary research on the "tobacco strategy," we hypothesized that bibliometric and text mining analyses would confirm that MQ functioned as an echo chamber for Pioneer-funded scientists.  This hypothesis was not supported.  Instead, mainstream psychology journals served as the echo chambers for race science.

@@

Figure \@ref(fig:wos) shows that, while MQ is among the "Pioneer-publishing" journals, a number of mainstream journals are more prominent: *Personality and Individual Differences* (PID), *Intelligence* (I), *Behavior Genetics* (BG), and *Psychological Reports* (PR).  In addition, only psychologist Richard Lynn appears to have published heavily in MQ.  Lynn became an assistant editor of MQ in 1979 (vol. XX, Nos. 1 & 2) and is listed as editor-in-chief on MQ's current website as of 2023-07-21.  He has also been president of Pioneer since the death of psychologist J. Phillippe Rushton in 2012 \cite{BeirichPioneerFundAssets2013}.  By contrast, a number of Pioneer-funded researchers have published in PID, I, and to a lesser degree BG: Bouchard, Eysenck, Jensen, Rushton, Vernon, and also Lynn.

@@

*Intelligence* (I) was founded in 1977, with psychologist Douglas Detterman as editor-in-chief from the founding until 2016.  Lloyd Humphreys was on the editorial board starting from 1977; by 1990 he had been joined by Jensen and Philip Vernon.  Richard Lynn joined the editorial board sometime between 1998 and 2002.  (Archive copies of the I editorial board page are not available from the journal's website from 1999 through 2001.)  I has been criticized for including Lynn and Gerhard Meisenberg — who was editor-in-chief of MQ in 2015-18 — on its editorial board until 2018 \cite{SainiSuperiorReturnRace2019}.  I is published by Elsevier.

@@

The fact that Pioneer-funded researchers published heavily in two mainstream psychology journals does not tell us anything about the content of their publications or the claims of their research.  We assembled a full-text corpus of articles published in MQ, PID, I, BG, PR, and Behavior and Brain Sciences (BBS) between 1960-2010 and used topic modeling to identify topics discussing race, intelligence, and both.

@@

Figure \@ref(fig:focal) and this discussion focuses on three topics identified in one of the 24 fitted topic models; see figures \@ref(fig:gamma-lg), \@ref(fig:gamma-md), \@ref(fig:gamma-sm) and supplemental visualizations S2-S4 for all topics from all models and S5-S7 for selected topics from all models.  In the focal model, topic 07 is strongly associated with MQ:  MQ published dozens of articles in this topic each year, and no other journal ever published more than a handful.  The Silge plot of the top 15 terms \cite{SilgeTopicModeling2017} contains racial terms (`races`, `whites`, `negroes` and potentially `europe`, `africa`, `india`, and `japan`) as well as `book`, likely reflecting the fact that MQ published a number of book reviews, while the psychology journals either did not or these were not available for the corpus.

Topic 22 is strongly associated with I and PID in the same way that 07 is associated with MQ.  The Silge plot does contain `jensen`, as well as a reference to Raymond Cattell, who played a major role in the development of factor analysis and intelligence testing but also advocated for eugenics, fascism, and Nazi race science \cite{MehlerBeyondismRaymondCattell1997}. However, the other authors named in this topic — John Horn and Peter Bentler — do not appear to have contributed to scientific racism.  Instead this topic appears to identify "mainstream" (non-race science) intelligence research, especially factor analysis and the debate over whether intelligence in unidimensional or multidimensional.  (Three other topics only associated with mainstream intelligence journals and terms were also identified by this topic model.)  This topic indicates that the model is not simply lumping race science research together with other intelligence research.

@@

In almost all years, most papers in topic 24 were published in mainstream journals rather than MQ.  Jensen's "How Much Can We Boost IQ and Scholastic Achievement?" \cite{JensenHowMuchCan1969} was published in 1969 in Harvard Educational Review (not included in this study), and the time series indicates that, around this time, there was an increase in articles in topic 24 in both PS and MQ (the only two journals in our corpus that were active at the time).  MQ shows another sharp increase in the late 1970s, around the time Lynn was brought on as assistant editor. This finding is consistent with the conclusion of a qualitative analysis that "Psychology became the dominant discipline upon the editorship of Richard Lynn and Gerhard Meisenberg in MQ, where some issues were almost entirely represented by IQ test–based studies" \cite[p. 110]{AdamsMisAppropriationBiological2021}.  PID published multiple papers in topic 24 almost immediately after it was founded, with I showing a more gradual increase between the mid-1970s and mid-1990s.

@@

It's notable that BG published very few articles in any race-and-intelligence topic identified by any of the 24 topic models. Among both academics and the general public, the field of behavior genetics is strongly associated with race science, and specifically race-and-intelligence research. Panofsky \cite{PanofskyMisbehavingScienceControversy2014} argues that, prior to Jensen's 1969 paper, behavior genetics emphasized the study of non-human animals and intentionally avoided associations with eugenics and public controversy more generally. In response to Jensen, critics such as Lewontin offered broad critiques of behavior genetics as such, and behavior geneticists in turn adopted a radical conception of academic freedom (without any sense of responsibility for the social implications of academic research) and a siege or wartime mentality, as demonstrated by Sandra Scarr's 1986 presidential address to the Behavior Genetics Association (BGA) \cite{ScarrThreeCheersBehavior1987}.  Scarr's address coincided with the period between 1970-1990 when BG published articles in topic 24 (including Scarr's address itself).

However, the topic model analysis suggests that by the 1990s, behavior genetics may have distanced itself somewhat from race science, albeit without directly repudiating it. This interpretation is further supported by two disciplinary histories of behavior genetics, one published in McGue in 2010 \cite{McGueEndBehavioralGenetics2010} and the other by Plomin in 2023 \cite{PlominCelebratingCenturyResearch2023}. In addition, the 1995 BGA presidential address by Glayde Whitney --- in which he criticized the "Marixist-Lysenkoist denial of genetics" and proposed that differences in murder rates between countries and cities was caused by racial genetic differences in intelligence, empathy, aggression, and impulsivity --- was published in MQ rather than BG. As a discipline, behavior genetics may have been prominent in forming an echo chamber for race science in the past, but appears to have been less receptive to such ideas in more recent history.

Instead, our results suggest that psychology has been the dominant discipline in which the race science echo chamber has been maintained. This may be the result of racial essentialization being maintained more easily in a discipline like psychology than in other disciplines that contribute to behavior genetics, like genetics and anthropology.  The typical practitioners in psychology do not have the degree of formal education about, and therefore knowledge of, relevant underlying genetic and anthropological research that has led many professional organizations in genetics and anthropology to formally reject race as a biologically meaningful concept \cite{anthropology1996aapa, american2018ashg, board1998aaa, FuentesAAPAStatementRace2019, nhgri2023, rotimi2022ashg, wynshawboris2020ashg}, undermining the core assumption of race science. As described by Winston, however, "[s]ince the 1990s, scientific racism has evolved in ways that sidestep or blunt the charge of essentialist categorization" \cite[p. 427]{WinstonWhyMainstreamResearch2020}, such as shifting the rhetoric away from race to concepts such as "national IQ" that are deployed to preserve an essentialist categorization of different human populations. Professional psychological organizations routinely publish statements rejecting racism or acknowledging psychology's complicity in systemic racism \cite{american2021apology, aps2021statement}, but stop short of rejecting race as a meaningful biological categorization of people, allowing for race science research to be smuggled into psychology and giving the appearance of continued scientific legitimacy.

@@

Using bibliometric and text-mining methods, this study finds that Pioneer-funded researchers published heavily in certain mainstream psychology journals, much more than in *Mankind Quarterly*; and that a distinct topic of race science discourse, centered on race and intelligence, can be identified in these same journals across the period 1960-2010.  These findings indicate that *Mankind Quarterly* (MQ) was less important as an echo chamber for late 20th century race science than mainstream psychology journals, especially *Intelligence* (I) and *Personality and Individual Differences* (PID).  Indeed, MQ published a very different kind of race science from the race-and-intelligence research published in I and PID.

We stress that we are not identifying any particular scientist as a racist, scientific racist, or race scientist based on the topic model results alone.  Scientific research can be utilized to promote scientific racism — and thus count as race science — independently of the views and attitudes of the scientist who conducted that research \cite{TaberyWhyStudyingGenetics2015, GillbornSoftlySoftlyGenetics2016, CarlsonQuantifyingContextualizingImpact2020, HennWhyDNANo2021}.  But this does not mean that scientists are not responsible for the downstream social effects of their 
research, even when these effects involve deliberate misinterpretation \cite{DouglasSciencePolicyValuefree2009, KitcherArgumentFreeInquiry1997, KouranyShouldKnowledgeBe2016, BlockIQHeritabilityInequality1974a}.  Like other citizens, scientists are responsible for mitigating the reasonably foreseen harms that result from their actions.  And we now have decades of empirical evidence that research on the genetics of intelligence — even research that does not itself touch on race — can and will be used to promote scientific racism and indeed racial violence \cite{MeyerWrestlingSocialBehavioral2023}.  For instance, while our findings suggest that behavior genetics has attempted to distance itself from scientific racism, it has done little to explicitly repudiate this research, especially in contrast to statements by biological anthropologists \cite{FuentesAAPAStatementRace2019} and human geneticists \cite{JacksonFacingOurHistory2023}.  Behavior genetics research — and research on human genetics more generally — may be especially susceptible to appropriation by scientific racists, which only strengthens the obligations these researchers have to prevent their research from causing harm \cite{CarlsonQuantifyingContextualizingImpact2020, OgbunugaforDNABasketballBirthday2022}.

Our distinction between race science and race science discourse highlights an important "engagement dilemma" for scientists challenging race science and scientific racism.  While the "tobacco strategy" for manufacturing doubt targets mainstream scientists and promotes fringe scientific views, its primary audiences are lawmakers and the general public.  The aim of the strategy is not to change the opinion of mainstream scientists, but instead to cultivate the impression that there is legitimate scientific uncertainty about the issue, reasonable disagreement among qualified experts \cite{FernandezPintoKnowBetterNot2017}.  On the one hand, refusing to engage with fringe scientists — or, often, pseudoscientists \cite{BhakthavatsalamVirtueEpistemologicalApproach2021} — means that false or misleading statements go unchallenged.  On the other hand, detailed, technical criticism legitimizes the fringe views, creating exactly the impression of reasonable scientific disagreement that the tobacco strategy is intended to produce.

In the case of race science, mainstream psychologists have often taken a non-engagement approach — whether due to a strong conception of academic freedom or a desire to avoid stoking controversy — with the result that race scientists have been able to present themselves as the representatives of "mainstream science on intelligence."  On the other hand, critics — most prominently Gould, Kamin, and Lewontin — have offered sophisticated technical critiques.  But these tend to spiral off into esoteric debates that are inaccessible to the general public, as with "Lewontin's fallacy" \cite{RosemanLewontinDidNot2021}.  Consequently these critiques are likely to be either ignored by the general public or taken as evidence of legitimate scientific debate.

@@

For $k=5, 10$ the distributions are quite noisy, difficult to interpret, and generally don't form very strong visual clusters.  As $k$ increases from 20 through 40, coherent bands start to appear for journal-distinctive topics in MQ, BG, I and, to a lesser degree, PID.  Note that the distinctive topics for MQ and I do not obviously coincide.  Above about $k=50$, the models appear to identify more fine-grained topics and the coherent bands fade.

@@

The 3 non-consensus documents reflected essential ambiguity in the operationalization of race science discourse.  One is a table of contents from a 1980 issue of I; arguably this document should have been excluded as irrelevant front matter, but arguably it legitimizes the research published in the issue.  The second case is a paper on racial differences in emotional intelligence that was classified as race science discourse by one author, while the other author felt it was more focused on psychometrics \cite{GignacGroupDifferencesEI2010}.  A finally, a 1990 article in MQ (<https://www.unz.com/print/MankindQuarterly-1990q3-00108/>) proposed to offer an economic explanation for crime.  While it has a few pages connecting race/ethnicity to both crime and poverty, it ultimately focuses on unemployment and alcohol sales, rather than race, as the key predictors of crime.

@@

A key limitation of the current study stems from the inability of text mining techniques to draw on a broader cultural context than what is represented in the corpus.  Scholars of racism and US race relations have noted a shift in the way racist attitudes are understood and expressed, from overt and direct to a subtle "color-blind" racism \cite{bonillasilva2022racism}. Color-blind racism represents a change in the primary mechanisms by which white privilege is maintained, using non-racialized language in furtherance of the racial status quo both institutionally and individually, and increasingly reliant on the invisibility of socio-cultural mechanisms to preserve inequalities. Color-blind racism involves the use of racialized "codes" or "dogwhistles," language that is superficially non-racial but carries racial implications, such as (in the US context) "violent inner city criminals" or "welfare queen" \cite{SaulDogwhistlesPoliticalManipulation2018}.  Alderfer notes that the racist implications of *The Bell Curve* only emerge gradually, with the opening of the book using the color-blind language of "group differences" \cite{AlderferScienceNonsciencePsychologists2003}.  Panofsky et al. show how the term "human biodiversity," originally developed for anti-racist purposes by biological anthropologist Jonathan Marks, has been inverted by white supremacists and is now a dogwhistle for biological racial hierarchy \cite{PanofskyHowWhiteNationalists2021}.  These cases suggest further that color-blindness can facilitate white supremacist appropriation of non-racial research — even contrary to the researchers' own intentions — as scientific racist readers re-interpret racial-neutral language as dogwhistles \cite{GillbornSoftlySoftlyGenetics2016, WillsAreClustersRaces2017}.

>>>>

acknowledgements: |
  | Thanks to Emily Merchant and John Jackson for some initial discussions that helped clarify the scope of the project and identify some key resources related to the Pioneer Fund and *Mankind Quarterly*.  Thanks to Anthony Sainez for help retrieving the *Mankind Quarterly* articles.  Thanks to Derek Devnich and James Dooley for their work attempting to secure the APA-published articles.  Thanks to Kevin Bird, John Jackson, and Manuela Fernández Pinot for comments on a draft of this paper. 
  | **Funding:** EL's work on this project was partially supported by UC Merced.  DH's work on this project was not supported by any specific funding. 
  | **Author contributions:** DJH (conceptualization, data curation, formal analysis, investigation, methodology, project administration, software, validation, visualization, writing-original draft, writing-review & editing); EL (conceptualization, methodology, software, validation, visualization, writing-original draft, writing-review & editing)
  | **Competing interests:** The authors have no competing interests to declare. 
  | **Data and materials availability:** Web of Science search results and article fulltext are not publicly available due to intellectual property restrictions.  Preprocessed analysis data (document-term counts and article metadata) and analysis scripts are available at *[doi]*. 
# bibliography: "references.bib"
output: 
  # https://github.com/rstudio/rticles/pull/486
  # skeleton: <https://github.com/christopherkenny/rticles/tree/master/inst/rmarkdown/templates/science>
  bookdown::pdf_book:
    base_format: rticles::science_article
    keep_tex: true
    move_figures: false
    draft: true
    number_sections: true
    latex_engine: 'xelatex'
header-includes:
  - \usepackage{lineno}
  - \linenumbers
---

@@

In the landmark case *Brown v Board of Education of Topeka* (1954), the US Supreme Court banned *de jure* educational segregation.  The Court's decision relied on expert testimony from psychologists and education researchers; but the segregationists also put forward their own experts, including Henry Garrett \cite{WinstonScienceServiceFar1998, JacksonScienceSegregationRace2005, SchafferScientificRacismAgain2007}.  As part of this segregationist reaction, MQ was created to provide a favorable venue for race scientists to publish their views, on the grounds that an "equalitarian dogma" created a censorious "taboo" against their research in mainstream publications \cite{TuckerFundingScientificRacism2002, JacksonMythicalTabooRace2020}.

In light of the scientific racist origins of Pioneer and MQ, and the significant attention MQ has received in recent historiography, we hypothesized that bibliometric and text mining analyses would show race science ideas originating in MQ, and from there being disseminated to mainstream publications.  This hypothesis was not supported.  Instead, mainstream psychology journals provided a venue for a distinct form of race science, based on intelligence research and originating outside of MQ.

@@

Figure \@ref(fig:wos) shows that, while MQ is among the "Pioneer-publishing" journals, a number of mainstream journals are more prominent: *Personality and Individual Differences* (PID), *Intelligence* (Int), *Behavior Genetics* (BG), and *Psychological Reports* (PR).  In addition, only psychologist Richard Lynn appears to have published heavily in MQ.  Lynn became an assistant editor of MQ in 1979 (vol. XX, Nos. 1 & 2) and is listed as editor-in-chief on MQ's current website as of 2023-07-21.  He has also been president of Pioneer since the death of psychologist J. Phillippe Rushton in 2012 \cite{BeirichPioneerFundAssets2013}.  By contrast, a number of Pioneer-funded researchers have published in PID, Int, and to a lesser degree BG: Bouchard, Eysenck, Jensen, Rushton, Vernon, and also Lynn.

@@

*Intelligence* (Int) was founded in 1977, with psychologist Douglas Detterman as editor-in-chief from the founding until 2016.  Lloyd Humphreys was on the editorial board starting from 1977; by 1990 he had been joined by Jensen and Philip Vernon.  Richard Lynn joined the editorial board sometime between 1998 and 2002.  (Archive copies of the Int editorial board page are not available from the journal's website from 1999 through 2001.)  Int has been criticized for including Lynn and Gerhard Meisenberg — who was editor-in-chief of MQ in 2015-18 — on its editorial board until 2018 \cite{SainiSuperiorReturnRace2019}.  Int is published by Elsevier.

@@

The fact that Pioneer-funded researchers published heavily in two mainstream psychology journals does not tell us anything about the content of their publications or the claims of their research.  We assembled a full-text corpus of articles published in MQ, PID, Int, BG, PR, and Behavior and Brain Sciences (BBS) between 1960-2010 and used topic modeling to identify topics discussing race, intelligence, and both.

@@

Figure \@ref(fig:focal) and this discussion focuses on three topics identified in one of the 24 fitted topic models; see figures \@ref(fig:gamma-lg), \@ref(fig:gamma-md), \@ref(fig:gamma-sm) and supplemental visualizations S2-S4 for all topics from all models and S5-S7 for selected topics from all models.  In the focal model, topic 07 is strongly associated with MQ:  MQ published dozens of articles in this topic each year, and no other journal ever published more than a handful.  The "Silge plot," showing the top 15 terms in the topic \cite{SilgeTopicModeling2017}, contains racial terms (`races`, `whites`, `negroes` and potentially `europe`, `africa`, `india`, and `japan`) as well as `book`, likely reflecting the fact that MQ published a number of book reviews, while the psychology journals either did not or these were not available for the corpus.

Topic 22 is strongly associated with Int and PID in the same way that 07 is associated with MQ.  The Silge plot does contain `jensen`, as well as a reference to Raymond Cattell, who played a major role in the development of factor analysis and intelligence testing but also advocated for eugenics, fascism, and Nazi race science \cite{MehlerBeyondismRaymondCattell1997}. However, the other authors named in this topic — John Horn and Peter Bentler — do not appear to have contributed to scientific racism.  Instead this topic appears to identify "mainstream" (non-race science) intelligence research, especially factor analysis and the debate over whether intelligence is unidimensional or multidimensional.  (Three other topics only associated with mainstream intelligence journals and terms were also identified by this topic model.)  This topic indicates that the model is not simply lumping race science research together with other intelligence research.

@@

In almost all years, most papers in topic 24 were published in mainstream journals rather than MQ.  Jensen's "How Much Can We Boost IQ and Scholastic Achievement?" \cite{JensenHowMuchCan1969} was published in 1969 in *Harvard Educational Review* (not included in this study), and the time series indicates that, in the early 1970s, there was an increase in articles in topic 24 in both PS and MQ (the only two journals in our corpus that were active at the time).  MQ shows another sharp increase in the late 1980s; using content analysis, Adams and Pilloud found that psychology was the dominant discipline in MQ in the period 1992-2018 \cite{AdamsMisAppropriationBiological2021}.  PID published multiple papers in topic 24 almost immediately after it was founded, with Int showing a more gradual increase between the mid-1970s and mid-1990s.

@@

It's notable that BG published very few articles in any race-and-intelligence topic identified by any of the 24 topic models. Among both academics and the general public, the field of behavior genetics is strongly associated with race science, and specifically race-and-intelligence research. Panofsky \cite{PanofskyMisbehavingScienceControversy2014} argues that, prior to Jensen's 1969 paper, behavior genetics emphasized the study of non-human animals and intentionally avoided associations with eugenics and public controversy more generally. In response to Jensen, critics such as Lewontin offered broad critiques of behavior genetics as such, and behavior geneticists in turn adopted a radical conception of academic freedom (without any sense of responsibility for the social implications of academic research) and a siege or wartime mentality, as illustrated by Sandra Scarr's 1986 presidential address to the Behavior Genetics Association (BGA) \cite{ScarrThreeCheersBehavior1987}.  Scarr's address coincided with the period between 1970-1990 when BG published articles in topic 24 (including Scarr's address itself).

However, the topic model analysis suggests that by the 1990s, behavior genetics may have distanced itself somewhat from race science, albeit without directly repudiating it. This interpretation is further supported by two disciplinary histories of behavior genetics published in BG in 2010 and 2023 \cite{McGueEndBehavioralGenetics2010, PlominCelebratingCenturyResearch2023}. In addition, the 1995 BGA presidential address by Glayde Whitney --- in which he criticized the "Marxist-Lysenkoist denial of genetics" and proposed that differences in murder rates between countries and cities was caused by racial genetic differences in intelligence, empathy, aggression, and impulsivity --- was published in MQ rather than BG. As a subdiscipline of psychology, behavior genetics may have been prominent in promoting race science in the past, but appears to have been less receptive to such ideas in more recent history.

Instead, our results suggest that other subdisciplines of psychology have provided the primary venue for mainstreaming race science.  Over the past few decades, many professional organizations in genetics and anthropology have made formal statements rejecting race as a biologically meaningful concept \cite{anthropology1996aapa, american2018ashg, board1998aaa, committee2020asa, FuentesAAPAStatementRace2019, nhgri2023, rotimi2022ashg, wynshawboris2020ashg}, undermining the core assumption of race science.  The American Psychological Association (APA) --- whose members are predominantly clinical psychologists --- and the Federation of Associations in Behavioral & Brain Sciences (FABBS) --- an organization whose purpose is to provide policy recommendations --- have openly rejected race as biologically meaningful \cite{american2021apology, baron2022fabbs}.  But research-focused psychological organizations like the Association for Psychological Science (APS) and Psychonomic Society (PS) stop short of rejecting race as biologically meaningful when they denounce racism \cite{aps2021statement, board2020psychonomic}.  Psychology as a discipline maintains space for several mythological race science narratives, such as allegations of a taboo against race and intelligence research \cite{jackson2021mythical}, claims that scientists engaging in or calling for any kind of anti-racism in the field is an ideological corruption of dispassionate and value-neutral science \cite{roberts_2022}, or arguments that holding race science to the same evidentiary standards as other psychological research is a violation of academic freedom \cite{herbert2023academic}.  And historically several race scientists have secured important gatekeeping positions within the scientific community, such as on the editorial boards of Int and PID.

@@

Using bibliometric and text-mining methods, this study finds that Pioneer-funded researchers published heavily in certain mainstream psychology journals, much more than in *Mankind Quarterly*; and that a distinct topic of race science discourse, centered on race and intelligence, can be identified in these same mainstream journals across the period 1960-2010.  These findings indicate that *Mankind Quarterly* (MQ) was less important as venue for late 20th century race science than mainstream psychology journals, especially *Intelligence* (Int) and *Personality and Individual Differences* (PID).  Indeed, until the 1990s, MQ published a very different kind of race science from the race-and-intelligence research published in Int and PID.  During this same period of time, race scientists served in prominent and influential positions in behavior genetics and psychology, including as society presidents and members of journal editorial boards.  In some cases, some race scientists simultaneously maintained active connections to both the mainstream scientific community and white supremacist organizations \cite{WinstonScientificRacismNorth2020, JacksonJr.ArthurJensenEvolutionary2022, SlobodianUnequalMindHow2023}.  As far as we are aware, no organization for scientific research in psychology has apologized for the historical contributions of their fields to scientific racism.

We are not identifying any particular scientist as a racist, scientific racist, or race scientist based on the topic model results alone.  Such claims require an analysis of documentary evidence that goes beyond the scope of the current study \cite{JacksonJr.ArthurJensenEvolutionary2022}.  We do assume that our readers, like us, regard white supremacy and scientific racism as morally odious and beyond the scope of reasonable debate \cite{SchroederLimitsDemocratizingScience2022}.  However, scientific research can be appropriated to promote scientific racism — and thus count as race science — even when this is contrary to the intentions of the original researchers themselves \cite{TaberyWhyStudyingGenetics2015, GillbornSoftlySoftlyGenetics2016, CarlsonQuantifyingContextualizingImpact2020, HennWhyDNANo2021}.  Panofsky et al. show how the term "human biodiversity," originally developed for anti-racist purposes by biological anthropologist Jonathan Marks, has been inverted by white supremacists and is now a dogwhistle for biological racial hierarchy \cite{PanofskyHowWhiteNationalists2021}.

Research on the genetics of intelligence has been socially harmful, not just in the somewhat abstract sense of promoting racial stigma, but in the concrete sense of being used to rationalize mass shootings and other acts of racial violence \cite{PronczukRacistResearcherExposed2022, MeyerWrestlingSocialBehavioral2023}.  While scientists do not fully control the downstream social effects of their research, like other citizens scientists are responsible for mitigating reasonably foreseeable harms that result from their actions \cite{BlockIQHeritabilityInequality1974a, DouglasSciencePolicyValuefree2009, KitcherArgumentFreeInquiry1997, KouranyShouldKnowledgeBe2016}.  These actions include not only individual decisions to research this topic or that, but also collective decisions about who is awarded positions of power or influence.  Given the way research on intelligence and behavior genetics has been used historically, these fields may be especially susceptible to appropriation by scientific racists.  This only strengthens the obligations of researchers in these areas to prevent their research from causing harm \cite{CarlsonQuantifyingContextualizingImpact2020, OgbunugaforDNABasketballBirthday2022}.

How should psychologists exercise these responsibilities?  We believe biological anthropology could provide an important model for psychology.  Anthropologists not only denounce racism in general, but the specific ways that anthropology has (and, in some cases, continues to be) used to provide cover for scientific racism.  Anthropologists produce both moral and technical critiques of race science, in venues and language that are appropriate for the general public.  And they recognize the importance of reparations, taking active steps to repair relationships with communities that have been harmed by anthropology.

@@

For $k=5, 10$ the distributions are quite noisy, difficult to interpret, and generally don't form very strong visual clusters.  As $k$ increases from 20 through 40, coherent bands start to appear for journal-distinctive topics in MQ, BG, Int and, to a lesser degree, PID.  Note that the distinctive topics for MQ and Int do not obviously coincide.  Above about $k=50$, the models appear to identify more fine-grained topics and the coherent bands fade.

@@

The 3 non-consensus documents reflected essential ambiguity in the operationalization of race science discourse.  One is a table of contents from a 1980 issue of Int; arguably this document should have been excluded as irrelevant front matter, but arguably it legitimizes the research published in the issue.  The second case is a paper on racial differences in emotional intelligence that was classified as race science discourse by one author, while the other author felt it was more focused on psychometrics \cite{GignacGroupDifferencesEI2010}.  A finally, a 1990 article in MQ (<https://www.unz.com/print/MankindQuarterly-1990q3-00108/>) proposed to offer an economic explanation for crime.  While it has a few pages connecting race/ethnicity to both crime and poverty, it ultimately focuses on unemployment and alcohol sales, rather than race, as the key predictors of crime.

@@

A key limitation of the current study stems from the inability of text mining techniques to draw on a broader cultural context than what is represented in the corpus.  Scholars of racism and US race relations have noted a shift in the way racist attitudes are understood and expressed, from overt and direct to a subtle "color-blind" racism \cite{Bonilla-SilvaRacismRacistsColorBlind2006}. Color-blind racism represents a change in the primary mechanisms by which white privilege is maintained, using non-racialized language in furtherance of the racial status quo both institutionally and individually, and increasingly reliant on the invisibility of socio-cultural mechanisms to preserve inequalities. Color-blind racism involves the use of racialized "codes" or "dogwhistles," language that is superficially non-racial but carries racial implications, such as (in the US context) "violent inner city criminals" or "welfare queen" \cite{SaulDogwhistlesPoliticalManipulation2018}.  Alderfer notes that the racist implications of *The Bell Curve* only emerge gradually, with the opening of the book using the color-blind language of "group differences" \cite{AlderferScienceNonsciencePsychologists2003}.  These cases suggest further that color-blindness can facilitate white supremacist appropriation of non-racial research — even contrary to the researchers' own intentions — as scientific racist readers re-interpret racial-neutral language as dogwhistles \cite{GillbornSoftlySoftlyGenetics2016, WillsAreClustersRaces2017}.
